Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-05 doc/invt.tex
solve a baseline shift problem in metad.h
invt.tex: shift the lambda index by 1

\newcommand{\Err}{E}
\newcommand{\ii}{\mathrm{i}}

@@

%For a large system at temperatures of interest,
%the distribution, $p^*_i$, often tends to
%be localized,
%%
%and to explore the global shape
%of the PMF, $-\ln p^*_i$,
%it is often advantageous to perform
%biased sampling that targets
%a wider and smoother distribution, $p_i$.
%%
%%Here, we refer to simulations that target
%%a flat or nearly flat distribution
%%as entropic or multicanonical sampling.
%
%To do so, we introduce a bias potential, $V_i$,
The challenge of sampling a multimodal distribution, $p^*_i$,
with deep valley(s) can often be eased by
biased sampling targeting a smoother distribution, $p_i$.
%
%
%where the proportionality constant, $C_i$,
%is found from the normalization condition,
%$\sum_{i = 1}^n \pi_i = 1$,
%in the second step.
%
%In particular,
%to make $\pi_i$ flat,
%the bias potential $V_i$
%must coincide with $\ln p^*_i$,
%up to an additive constant.
%%
%In practice, $V_i$ may be non-optimal or contain error,
%and thus $\pi_i$ need not
%coincide with the intended distribution $p_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $p_i$ (i.e. $p_i \equiv 1/n$)
or nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
By introducing a bias potential, $V_i$,
the biased equilibrium distribution becomes
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p^*_i \, e^{-V_i}
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  \propto
  p^*_i \, e^{-V_i}
  ,
  \label{eq:pi_p_V}
\end{equation}
%
and our aim is to find a biased potential
such that the resulting $\pi_i$ would coincide
with the desired $p_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
and the PMF, $-\ln p^*_i$, can be readily 
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

Many methods\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre, huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
have been developed to find the proper bias potential.
%
In a class of equilibrium FDS methods\cite{
  mezei1987, berg1992, *lee1993, marinari1992, *lyubartsev1992},
the bias potential, $V_i$, is
estimated a priori and fixed
during simulation,
%
and a correction to the bias potential
is derived from the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

@@

%We will mainly focus on the adaptive methods below.
%
In the original WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
including the Gaussian updating scheme
used in metadynamics,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian of $i-j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%One can show that with a constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $p_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $V_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
%We will suggest a slight modification of Eq. \eqref{eq:alpha_invt}
%whereby the origin of $t$ is shifted by an amount
%determined from the initial error.
%
We will first express the error of
the bias potential, $V_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

@@

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $V_i(t)$,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation, $x_i(t)$, of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bar v(t)$,
for average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot V_i(t)
  \equiv
  \frac{ dV_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
$
h_i(t) = \delta_{i, i(t)}
$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.

@@

%The average growth rate of $v(t)$ is given by
%\begin{equation}
%  \frac{ d \bar v (t) }
%       { d t }
%  =
%  \alpha(t)
%  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
%  .
%\label{eq:dvbardt}
%\end{equation}
%

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get % a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_singlebin}
\end{equation}
%
where the shifted histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  \equiv
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$:
%i.e. we may write
\begin{equation}
  \langle \xi_i(t) \, \xi_i(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}

@@

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $x_i(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $u(q(t))$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $u(q(t))$ should be a constant:
%
%We will further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
%By the Euler-Lagrange equation, we get
%$$
%\begin{aligned}
%0
%&=
%\frac{d}{dt} \int_0^T
%  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
%\\
%&= \int_0^T
%  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
%\end{aligned}
%$$
%where we have dropped the boundary terms
%by assuming Eq. \eqref{eq:kappa_limit}
%and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
%In order for this to hold for any $t$,
%we must have
%%
%$
%\ddot u\bigl( q(t') \bigr) = 0
%,
%$
%or
\begin{equation}
  \dot u\bigl( q(t) \bigr) = c
  .
\label{eq:dudt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:u_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and the other constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

@@

>>>>

\newcommand{\Err}{E}
\newcommand{\ii}{\mathrm{i}}
%bin average
\newcommand{\bav}[1]{#1_\mathrm{av}}

@@

%For a large system at temperatures of interest,
%the distribution, $p^*_i$, often tends to
%be localized,
%%
%and to explore the global shape
%of the PMF, $-\ln p^*_i$,
%it is often advantageous to perform
%biased sampling that targets
%a wider and smoother distribution, $p_i$.
%%
%%Here, we refer to simulations that target
%%a flat or nearly flat distribution
%%as entropic or multicanonical sampling.
%
%To do so, we introduce a bias potential, $u_i$,
The challenge of sampling a multimodal distribution, $p^*_i$,
with deep valley(s) can often be eased by
biased sampling targeting a smoother distribution, $p_i$.
%
%
%where the proportionality constant, $C_i$,
%is found from the normalization condition,
%$\sum_{i = 1}^n \pi_i = 1$,
%in the second step.
%
%In particular,
%to make $\pi_i$ flat,
%the bias potential $u_i$
%must coincide with $\ln p^*_i$,
%up to an additive constant.
%%
%In practice, $u_i$ may be non-optimal or contain error,
%and thus $\pi_i$ need not
%coincide with the intended distribution $p_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $p_i$ (i.e. $p_i \equiv 1/n$)
or nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
By introducing a bias potential, $u_i$,
the biased equilibrium distribution becomes
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p^*_i \, e^{-u_i}
  =
  \frac{                p^*_i \, e^{-u_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-u_j} }
  \propto
  p^*_i \, e^{-u_i}
  ,
  \label{eq:pi_p_V}
\end{equation}
%
and our aim is to find a biased potential
such that the resulting $\pi_i$ would coincide
with the desired $p_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  u_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
and the PMF, $-\ln p^*_i$, can be readily
deduced from the bias potential. % $u_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $u_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

Many methods\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre, huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
have been developed to find the proper bias potential.
%
In a class of equilibrium FDS methods\cite{
  mezei1987, berg1992, *lee1993, marinari1992, *lyubartsev1992},
the bias potential, $u_i$, is
estimated a priori and fixed
during simulation,
%
and a correction to the bias potential
is derived from the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $u_i$ as
%
\begin{equation}
  u^\mathrm{corr}_i
  =
  u_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-u_i},
    \\
    p_i
    &\propto p^*_i \, e^{-v^{0, \mathrm{corr}}_i}.
  \end{aligned}
  $$
}
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

@@

%We will mainly focus on the adaptive methods below.
%
In the original WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $u_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
including the Gaussian updating scheme
used in metadynamics,
several neighboring bins are updated:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian of $i-j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%One can show that with a constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $p_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $u_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
%%
%The source of the deviation is two-fold.
%%
%First, since $u_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
%We will suggest a slight modification of Eq. \eqref{eq:alpha_invt}
%whereby the origin of $t$ is shifted by an amount
%determined from the initial error.
%
We will first express the error of
the bias potential, $u_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

@@

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  u_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $u_i(t)$,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation, $x_i(t)$, of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bav{v}(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bav{v}(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
$
h_i(t) = \delta_{i, i(t)}
$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.

@@

%The average growth rate of $v(t)$ is given by
%\begin{equation}
%  \frac{ d \bav{v} (t) }
%       { d t }
%  =
%  \alpha(t)
%  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
%  .
%\label{eq:dvbardt}
%\end{equation}
%

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get % a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_singlebin}
\end{equation}
%
where the shifted histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  \equiv
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$:
%i.e. we may write
\begin{equation}
  \langle \xi_i(t) \, \xi_i(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \bav{\dot v}
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot y\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
y(q')
&\equiv
e^{q' - q(T)}.
\label{eq:y_def}
\end{align}

@@

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot y}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $x_i(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
%We will further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot y(t') \, \dot y(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot y\bigl( q(t') \bigr) \,
%  \dot y\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
%By the Euler-Lagrange equation, we get
%$$
%\begin{aligned}
%0
%&=
%\frac{d}{dt} \int_0^T
%  \dot y\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
%\\
%&= \int_0^T
%  \ddot y\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
%\end{aligned}
%$$
%where we have dropped the boundary terms
%by assuming Eq. \eqref{eq:kappa_limit}
%and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t y^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
%In order for this to hold for any $t$,
%we must have
%%
%$
%\ddot y\bigl( q(t') \bigr) = 0
%,
%$
%or
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and the other constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

@@

\subsection{\label{sec:lj}Lennard-Jones}
